Are translation equivalents special? Evidence from simulations and empirical data from bilingual infants

### 2021-12-07 paper/paper_TE_bilingual_vocabulary_model.Rmd
Updated paper

[^1]: We note that in some cases a singlet may not have a close translation equivalent, for example the recently popular Danish word “hygge” refers to a type of cozy contentment and enjoyment of simplicity. Moreover, even when a word that could be considered a translation equivalent exists, adults’ concepts may not be congruent across languages (e.g., Malt & Majid, 2013). However, in early development, most vocabulary words are reasonably concrete, and evidence suggests that concepts are largely shared across languages (Storm et al., 2020; White et al., 2020). Thus, we make the assumption that most early words have a translation equivalent and that these are fairly congruent for the age range of focus in the current paper.

@@

## 2.2.3 Testing Prediction Set 2: The vocabulary composition of bilingual children at different developmental levels
Prediction Set 2 pertained to expected patterns of acquisition of translation equivalents and singlets for children of different developmental levels. In our data set, developmental level was approximated by children’s age. Figure 3 Panel B shows the concept vocabulary (CONCEPT) of the bilingual children as a function of different ages (a proxy for developmental level), used to estimate the number of LEARNABLE words. To illustrate the acquisition of translation equivalents and singlets at different developmental levels, we divided children into three age groups: younger children of 18–22 months, middle children of 23–27 months, and older children of 28–33 months.

@@

# testing prediction 2a: To compare CONCEPT among the 3 LEARNABLE groups in the stacked bar chart

## correlation between age (in months) and CONCEPT
cor_age_CONCEPT <- tidy(cor.test(keepers_ws_TE$age_continuous, keepers_ws_TE$concept_vocab, method = "pearson"))

## one-way ANOVE to compare CONCEPT among the 3 LEARNABLE groups
keepers_ws_TE_3LEARNABLEsubset <- keepers_ws_TE %>%
  # Subset developmental level (i.e., Average_90percentile) by 3 levels 
  mutate(LEARNABLE_subset = case_when(age_months_percentile <= 22 ~ "18-22 months (V = 244.9 - 451.9)",
                              age_months_percentile > 22 & age_months_percentile <= 27 ~ "23-27 months (V = 491.8 - 604.1)",
                              TRUE ~ "28-33 months (V = 620.4 - 638.9)"))

aov_age_CONCEPT <- tidy(aov(concept_vocab ~ LEARNABLE_subset, data = keepers_ws_TE_3LEARNABLEsubset))

pairwise.t.test(keepers_ws_TE_3LEARNABLEsubset$concept_vocab, keepers_ws_TE_3LEARNABLEsubset$LEARNABLE_subset,
                p.adjust.method = "BH")

## mean CONCEPT by LEARNABLE subset
mean_concept_LEARNABLEsubset <- keepers_ws_TE_3LEARNABLEsubset %>%
  group_by(LEARNABLE_subset) %>%
  summarise(mean_concept = mean(concept_vocab, na.rm = T),
            sd_concept = sd(concept_vocab, na.rm = T),
            min_concept = min(concept_vocab, na.rm = T),
            max_concept = max(concept_vocab, na.rm = T)) %>%
  mutate(across(is.numeric, ~ round(.,1)))

# testing prediction 2b: To compare TE among the 3 LEARNABLE groups in the stacked bar chart

## correlation between age (in months) and TE
cor_age_TE <- tidy(cor.test(keepers_ws_TE$age_continuous, keepers_ws_TE$number_of_te, method = "pearson"))

## one-way ANOVA to compare TE among LEARNABLE
stacked_byLEARNABLE_wide <- keepers_ws_TE %>%
  select(baby_id, age_months_percentile, balance_vocab, word_vocab, concept_vocab, 
         number_of_te, total_singlet_dom, total_singlet_nondom) %>%
  # Subset developmental level (i.e., Average_90percentile) by 3 levels 
  mutate(LEARNABLE_subset = case_when(age_months_percentile <= 22 ~ "18-22 months",
                              age_months_percentile > 22 & age_months_percentile <= 27 ~ "23-27 months",
                              TRUE ~ "28-33 months")) %>%
  # Subset balance level by 3 groups
  mutate(BalSubset = case_when(balance_vocab > 0.35 ~ "Most Balanced",
                               balance_vocab <= 0.20 ~ "Least Balanced",
                               TRUE ~ "Medium Balanced"))

aov_age_TE <- tidy(aov(number_of_te ~ LEARNABLE_subset, data = stacked_byLEARNABLE_wide))

pairwise.t.test(stacked_byLEARNABLE_wide$number_of_te, stacked_byLEARNABLE_wide$LEARNABLE_subset,
                 p.adjust.method = "BH")

## mean TE by LEARNABLE subset
mean_TE_LEARNABLEsubset <- stacked_byLEARNABLE_wide %>%  
  group_by(LEARNABLE_subset) %>%
  summarise(mean_TE = mean(number_of_te, na.rm=T), 
            sd_TE = sd(number_of_te, na.rm=T),
            min_TE = min(number_of_te, na.rm=T),
            max_TE = max(number_of_te, na.rm=T)) %>%
  mutate(across(is.numeric, ~ round(.,1)))

# testing prediction 2c: To compare DOM-SINGLET among the 3 LEARNABLE groups in the stacked bar chart

## correlation between DOM-SINGLET and age (in month)
cor_age_DOMSING <- tidy(cor.test(keepers_ws_TE$age_continuous, keepers_ws_TE$total_singlet_dom, method = "pearson"))

@@

## one-way ANOVA to compare DOM-SINGLET among LEARNABLE
aov_balance_DOMSING <- tidy(aov(total_singlet_dom ~ BalSubset, data = stacked_byLEARNABLE_wide))

pairwise.t.test(stacked_byLEARNABLE_wide$total_singlet_dom, stacked_byLEARNABLE_wide$BalSubset,
                 p.adjust.method = "BH")

## mean DOM-SINGLET by LEARNABLE subset
mean_DOMSINGLET_LEARNABLEsubset <- stacked_byLEARNABLE_wide %>%  
  group_by(BalSubset) %>%
  summarise(mean_DOM_SINGLET = mean(total_singlet_dom, na.rm=T), 
            sd_DOM_SINGLET = sd(total_singlet_dom, na.rm=T),
            min_DOM_SINGLET = min(total_singlet_dom, na.rm=T),
            max_DOM_SINGLET = max(total_singlet_dom, na.rm=T)) %>%
  mutate(across(is.numeric, ~ round(.,1)))

# testing Prediction 2d: To compare NONDOM-SINGLET among the 3 LEARNABLE groups in the stacked barchart

## correlation between NONDOM-SINGLET and age (in month)
cor_age_NONDOMSING <- tidy(cor.test(keepers_ws_TE$age_continuous, keepers_ws_TE$total_singlet_nondom, method = "pearson"))

@@

## one-way ANOVA to compare NONDOM-SINGLET among LEARNABLE
aov_balance_NONDOMSING <- tidy(aov(total_singlet_nondom ~ BalSubset, data = stacked_byLEARNABLE_wide))

pairwise.t.test(stacked_byLEARNABLE_wide$total_singlet_nondom, stacked_byLEARNABLE_wide$BalSubset,
                 p.adjust.method = "BH")

## mean NONDOM-SINGLET by LEARNABLE subset
mean_NONDOMSINGLET_LEARNABLEsubset <-stacked_byLEARNABLE_wide %>% 
  group_by(BalSubset) %>%
  summarise(mean_NONDOM_SINGLET = mean(total_singlet_nondom, na.rm=T), 
            sd_NONDOM_SINGLET = sd(total_singlet_nondom, na.rm=T),
            min_NONDOM_SINGLET = min(total_singlet_nondom, na.rm=T),
            max_NONDOM_SINGLET = max(total_singlet_nondom, na.rm=T)) %>%
  mutate(across(is.numeric, ~ round(.,1)))
```

Prediction 2a was that older children (i.e., those at a later developmental level) would have larger concept vocabularies than younger children (i.e., those at an earlier developmental level). We observed a positive correlation between age (used as a proxy for developmental level, which determines LEARNABLE) and concept vocabulary (CONCEPT) in our dataset, _r_(`r cor_age_CONCEPT$parameter[1]`) = `r round(cor_age_CONCEPT$estimate[1],2)`, $p `r papaja::printp(cor_age_CONCEPT$p.value[1])`$, and therefore confirmed the prediction. This pattern was further confirmed by a one-way ANOVA, where the three age groups significantly differed in the number of concept vocabulary they produced, ($F$(`r aov_age_CONCEPT$df[1]`, `r aov_age_CONCEPT$df[2]`) = `r round(aov_age_CONCEPT$statistic[1],2)`, $p `r papaja::printp(aov_age_CONCEPT$p.value[1])`$). Older children of 28–33 months (i.e., at a later developmental level) produced the most with an average concept vocabulary of `r round(mean_concept_LEARNABLEsubset$mean_concept[3],2)` ($p$s < .001), middle children of 23–27 months (i.e., at an intermediate developmental level) produced an average concept vocabulary of `r round(mean_concept_LEARNABLEsubset$mean_concept[2],2)`, and younger children of 18–22 months (i.e., at an earlier developmental level) produced the least with an average concept vocabulary of `r round(mean_concept_LEARNABLEsubset$mean_concept[1],2)` ($p$s < .001).

Prediction 2b was that older children would produce more translation equivalents than younger children. First, we observed a positive correlation between age (our proxy for LEARNABLE) and number of translation equivalents in our dataset, _r_(`r cor_age_TE$parameter[1]`) = `r round(cor_age_TE$estimate[1],2)`, $p `r papaja::printp(cor_age_TE$p.value[1])`$, and therefore confirmed the prediction. In a one-way ANOVA with age group as factor, we further found that groups differed in how many translation equivalents they produced ($F$(`r aov_age_TE$df[1]`, `r aov_age_TE$df[2]`) = `r round(aov_age_TE$statistic[1],2)`, $p `r papaja::printp(aov_age_TE$p.value[1])`$). Younger children of 18–22 months produced an average of `r round(mean_TE_LEARNABLEsubset$mean_TE[1],2)` translation equivalents, middle children of 23–27 months produced an average of `r round(mean_TE_LEARNABLEsubset$mean_TE[2],2)` translation equivalents, and older children of 28–33 months produced an average of `r round(mean_TE_LEARNABLEsubset$mean_TE[3],2)` translation equivalents ($p$s < .01).

Prediction 2c was that both older children and those with the least balanced vocabularies (BALANCE) would produce more dominant-language singlets (DOM-SINGLET). This pattern was confirmed by the results from our dataset, with a positive correlation between dominant-language singlets (DOM-SINGLET) and age (which determined LEARNABLE), _r_(`r cor_age_DOMSING$parameter[1]`) = `r round(cor_age_DOMSING$estimate[1],2)`, $p `r papaja::printp(cor_age_DOMSING$p.value[1])`$, and a negative correlation between BALANCE and dominant-language singlets (DOM-SINGLET), _r_(`r cor_BALANCE_DOMSING$parameter[1]`) = `r round(cor_BALANCE_DOMSING$estimate[1],2)`, $p `r papaja::printp(cor_BALANCE_DOMSING$p.value[1])`$. As shown in Figure 3 Panel B, children were divided into least balanced (range of balance: .00 - .20), medium balanced (range of balance: .20 - .35) and most balanced (range of balance: .35 - .50) groups (i.e., the same criteria as in Figure 5). In a one-way ANOVA with balance group as a between-subjects factor, we observed that the least balanced children produced the most singlets in their dominant language ($p$s < .001), with the least balanced, medium balanced, most balanced children producing respectively: `r round(mean_DOMSINGLET_LEARNABLEsubset$mean_DOM_SINGLET[1],2)`, `r round(mean_DOMSINGLET_LEARNABLEsubset$mean_DOM_SINGLET[2],2)`, and `r round(mean_DOMSINGLET_LEARNABLEsubset$mean_DOM_SINGLET[3],2)` words in their dominant language ($F$(`r aov_balance_DOMSING$df[1]`, `r aov_balance_DOMSING$df[2]`) = `r round(aov_balance_DOMSING$statistic[1],2)`, $p `r papaja::printp(aov_balance_DOMSING$p.value[1])`$).

Prediction 2d was that older children and those with the most balanced vocabularies (BALANCE) would produce more singlets in their non-dominant language. This pattern was also observed in our dataset, with a positive correlation between the number of non-dominant singlets (NONDOM-SINGLET) and age (which determined LEARNABLE), _r_(`r cor_age_NONDOMSING$parameter[1]`) = `r round(cor_age_NONDOMSING$estimate[1],2)`, $p = `r papaja::printp(cor_age_NONDOMSING$p.value[1])`$, and a positive correlation between BALANCE and the number of non-dominant singlets (NONDOM-SINGLET), _r_(`r cor_BALANCE_NONDOMSING$parameter[1]`) = `r round(cor_BALANCE_NONDOMSING$estimate[1],2)`, $p `r papaja::printp(cor_BALANCE_NONDOMSING$p.value[1])`$. In a one-way ANOVA with balance group as a between-subjects factor, we confirmed that children who differed in how balanced their vocabulary knowledge was also differed in how many singlets they produced in their non-dominant language ($F$(`r aov_balance_NONDOMSING$df[1]`, `r aov_balance_NONDOMSING$df[2]`) = `r round(aov_balance_NONDOMSING$statistic[1],2)`, $p `r papaja::printp(aov_balance_NONDOMSING$p.value[1])`$). As shown in Figure 3 Panel B, we observed that children produced very few singlets in their non-dominant language, although the most balanced children produced the most singlets in their non-dominant language (mean of the most balanced children = `r round(mean_NONDOMSINGLET_LEARNABLEsubset$mean_NONDOM_SINGLET[3],2)` > mean of the medium balanced children = `r round(mean_NONDOMSINGLET_LEARNABLEsubset$mean_NONDOM_SINGLET[2],2)` > mean of the most balanced children = `r round(mean_NONDOMSINGLET_LEARNABLEsubset$mean_NONDOM_SINGLET[1],2)`; $p$s < .001).

## 2.2.4 Testing Prediction Set 3: Rate of translation equivalent learning
Prediction Set 3 pertained to the overall nature of translation equivalent learning, describing expected patterns of translation equivalent learning under the Neutral Account, the Avoidance Account, or the Preference Account. To directly test the correspondence of our data with these different accounts, we built a linear regression model predicting the observed number of translation equivalents from the Bilingual Vocabulary Model using the formula TE = DOM×NONDOM/LEARNABLE , and we allowed the model to estimate the BIAS parameter.

First, we will walk through the parameters in this model. The size of dominant vocabulary (DOM) and size of non-dominant vocabulary (NONDOM) were taken to be the number of words produced by individual children observed in the vocabulary data. The number of learnable words (LEARNABLE) was determined by averaging English and French productive CDI vocabulary at the 90th percentile at different ages (percentile information obtained from Wordbank; Frank et al., 2016). Table 7 lists the denominators derived from averaging these percentiles, or LEARNABLE, at different ages. For example, for an 18 month-old infant, the denominator was 244.9 words which was calculated by averaging the 268.7 English words and 221.1 French words, based on what 18-month-old children would typically produce at the 90th percentile. For children who were between 31 to 33 months in our dataset, the 90th percentile of 30-month-old children was used since 90th percentile information was available only up to 30 months.

```{r table7, results = "asis", cache = TRUE}
# read in the percentile file
ws_percentile <- read.csv(here::here("data_keepers/ws_percentile.csv")) %>%
  # filter out age ranges that is smaller than 18 months
  filter(age_months_percentile >= 18) %>%
  # as the upper age limit of the CDI-WS is 30m, create a new row for age 30-33m
  mutate(age_months_percentile = case_when(age_months_percentile == 30 ~ "30 - 33", 
                                           TRUE ~ as.character(age_months_percentile))) %>%
  # select only variables relevant for the 90th percentile
  select(age_months_percentile, EngWS_90percentile, FrWS_90percentile, Average_90percentile) %>%
  # rename variables for presentation
  rename("Age (months)" = age_months_percentile,
         "Number of English words produced at 90th percentile" = EngWS_90percentile,
         "Number of French words produced at 90th percentile" = FrWS_90percentile,
         "Average (LEARNABLE)" = Average_90percentile) %>%
  # round to 1 decimal place
  mutate(across(where(is.numeric),round,1))

kable(ws_percentile,
        format = "latex", booktabs = TRUE, align = "c",
        caption = "Table 7. The number of total English and French productive CDI vocabulary at the 90th percentile at different ages, and the average between the two which serves as the denominator in our computation model.") %>%
  kableExtra::kable_styling(font_size = 10) %>%
  column_spec(1, width = "75px") %>%
  column_spec(2:4, width = "125px")

@@

@@

## creating predicted TEs for simulation
keepers_ws_TE_simulation <- keepers_ws_TE %>%
  mutate(Predicted_TE_percentile = total_words_dom * total_words_nondom/Average_90percentile) %>%
  select(baby_id, ID_testdate, Predicted_TE_percentile, number_of_te, Average_90percentile, total_words_dom, total_words_nondom, 
         word_vocab, age_months_binned)

## linear regression model
test_prediction3_model <- keepers_ws_TE_simulation %>%
  lm(number_of_te*(Average_90percentile) ~ 0 + total_words_dom:total_words_nondom, data = .)

@@

Our model showed an excellent model fit of $R^2$ = `r round(summary(test_prediction3_model)$r.squared,2)`, indicating that our model explained `r round(summary(test_prediction3_model)$r.squared,2) * 100`% of the variance in bilinguals’ translation equivalent knowledge. The linear regression model estimated a BIAS coefficient of `r round(summary(test_prediction3_model)$coefficient[1],2)`, $p `r papaja::printp(summary(test_prediction3_model)$coefficient[4])`$. This value is extremely close to 1, suggesting that our data are consistent with the account whereby translation equivalents are learned equivalently to other words.

To illustrate the close fit between the Neutral Account and our data, we used the Bilingual Vocabulary Model formula TE = 1 × (DOM×NONDOM/LEARNABLE) to estimate each child’s expected translation equivalent knowledge (setting BIAS = 1), which is plotted against our observed data in Figure 6. Expected and observed translation equivalents were closely aligned with the Neutral Account of the Bilingual Vocabulary Model (i.e., BIAS = 1), suggesting that the Neutral Account provides a parsimonious explanation for bilinguals’ translation equivalent knowledge. This provides evidence for the notion that translation equivalents are neither harder nor easier to learn than singlets in bilingual vocabulary learning. Note that visual inspection suggested that there could be some possible outliers. Cook’s distance was estimated for our linear regression model listed above and identified `r words(length(number_of_outliers$ID_testdate))` data points with a cook’s distance over 0.4. After removing those `r words(length(number_of_outliers$ID_testdate))` data points, the linear regression model returned a coefficient of `r round(summary(model_remove_outliers)$coefficient[1],2)`, $p `r papaja::printp(summary(model_remove_outliers)$coefficient[4])`$, with $R^2$ = `r round(summary(model_remove_outliers)$r.squared,2)`. As the model fit was similar to the model without eliminating the `r words(length(number_of_outliers$ID_testdate))` outlier data points, we proceeded with the full data set keeping the `r words(length(number_of_outliers$ID_testdate))` potential outlier data points.

@@

Despite the good overall fit to the data, a close examination of Figure 6 suggested that the model might less closely fit the data of children with smaller vocabulary sizes. Figure 7 displays the model fit separately for children with a word vocabulary (WORD) less than 300 words and those with a word vocabulary (WORD) of 300 or greater. Based on visual inspection, the slope of translation equivalent learning appeared steeper for children with less than 300 total vocabulary, suggesting that translation equivalents are more easily learned (i.e., BIAS > 1); whereas the slope of translation equivalent learning appeared to align with the Neutral Account of the Bilingual Vocabulary Model (i.e., BIAS = 1) for children with more than 300 total vocabulary. To further explore this pattern, we ran the same linear regression twice, separately for children with less than 300 total vocabulary and for those with more than 300 total vocabulary. The model for those with larger total word vocabulary (WORD) returned a coefficient of BIAS = `r round(summary(percentile_model_more300w)$coefficient[1],2)`, $p `r papaja::printp(summary(percentile_model_more300w)$coefficient[4])`$, whereas the model for those with less than 300 total word vocabulary (WORD) returned a coefficient of BIAS = `r round(summary(percentile_model_less300w)$coefficient[1],2)`, $p `r papaja::printp(summary(percentile_model_less300w)$coefficient[4])`$. Both models fit well, although a somewhat better fit was obtained for children with larger vocabulary size ($R^2$ = `r round(summary(percentile_model_more300w)$r.squared,2)`) than children with smaller vocabulary size ($R^2$ = `r round(summary(percentile_model_less300w)$r.squared,2)`). Overall, this analysis suggests that translation equivalent learning for children with larger vocabularies corresponds best to the Neutral Account, but translation equivalent learning for children with smaller vocabularies corresponds best to the Preference Account.

>>>>

[^1]: We note that in some cases a singlet may not have a close translation equivalent, for example the recently popular Danish word _“hygge”_ refers to a type of cozy contentment and enjoyment of simplicity. Moreover, even when a word that could be considered a translation equivalent exists, adults’ concepts may not be congruent across languages (e.g., Malt & Majid, 2013). However, in early development, most vocabulary words are reasonably concrete, and evidence suggests that concepts are largely shared across languages (Storm et al., 2020; White et al., 2020). Thus, we make the assumption that most early words have a translation equivalent and that these are fairly congruent for the age range of focus in the current paper.

@@

## 2.2.3 Testing Prediction Set 2: The vocabulary composition of bilingual children at different developmental levels
Prediction Set 2 pertained to expected patterns of acquisition of translation equivalents and singlets for children of different developmental levels. In our data set, developmental level was approximated by children’s age. Figure 2 Panel B shows the number of translation equivalents (TE), singlets in the dominant language (DOM-SINGLET) and in the non-dominant language (NONDOM-SINGLET) of the bilingual children as a function of different ages (a proxy for developmental level), used to estimate the number of LEARNABLE words.

@@

# testing prediction 2a: Relationship between CONCEPT and Age
## correlation between age (in months) and CONCEPT
cor_age_CONCEPT <- tidy(cor.test(keepers_ws_TE$age_continuous, keepers_ws_TE$concept_vocab, method = "pearson"))

# testing prediction 2b: Relationship between TE and Age
## correlation between age (in months) and TE
cor_age_TE <- tidy(cor.test(keepers_ws_TE$age_continuous, keepers_ws_TE$number_of_te, method = "pearson"))

# testing prediction 2c: Relationship among DOM-SINGLET, Age, and BALANCE
## correlation between DOM-SINGLET and age (in month)
cor_age_DOMSING <- tidy(cor.test(keepers_ws_TE$age_continuous, keepers_ws_TE$total_singlet_dom, method = "pearson"))

@@

## mean DOM-SINGLET by BALANCE subset
mean_DOMSINGLET_BALANCEsubset <- keepers_ws_TE %>%
  mutate(bin_BALANCE = cut(balance_vocab, breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5), include.lowest=T,
                     labels = c("0.1", "0.2", "0.3", "0.4", "0.5"))) %>%
  group_by(bin_BALANCE) %>%
  summarise(mean_DOM_SINGLET = mean(total_singlet_dom, na.rm=T), 
            sd_DOM_SINGLET = sd(total_singlet_dom, na.rm=T),
            min_DOM_SINGLET = min(total_singlet_dom, na.rm=T),
            max_DOM_SINGLET = max(total_singlet_dom, na.rm=T)) %>%
  mutate(across(is.numeric, ~ round(.,0)))

# testing Prediction 2d: Relationship among NONDOM-SINGLET, Age, and BALANCE 
## correlation between NONDOM-SINGLET and age (in month)
cor_age_NONDOMSING <- tidy(cor.test(keepers_ws_TE$age_continuous, keepers_ws_TE$total_singlet_nondom, method = "pearson"))

@@

## mean DOM-SINGLET by BALANCE subset
mean_NONDOMSINGLET_BALANCEsubset <- keepers_ws_TE %>%
  mutate(bin_BALANCE = cut(balance_vocab, breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5), include.lowest=T,
                     labels = c("0.1", "0.2", "0.3", "0.4", "0.5"))) %>%
  group_by(bin_BALANCE) %>%
  summarise(mean_NONDOM_SINGLET = mean(total_singlet_nondom, na.rm=T), 
            sd_NONDOM_SINGLET = sd(total_singlet_nondom, na.rm=T),
            min_NONDOM_SINGLET = min(total_singlet_nondom, na.rm=T),
            max_NONDOM_SINGLET = max(total_singlet_nondom, na.rm=T)) %>%
  mutate(across(is.numeric, ~ round(.,0)))

@@

Prediction 2a was that older children (i.e., those at a later developmental level) would have larger concept vocabularies than younger children (i.e., those at an earlier developmental level). We observed a positive correlation between age (used as a proxy for developmental level, which determines LEARNABLE) and concept vocabulary (CONCEPT) in our dataset, _r_(`r cor_age_CONCEPT$parameter[1]`) = `r round(cor_age_CONCEPT$estimate[1],2)`, $p `r papaja::printp(cor_age_CONCEPT$p.value[1])`$, and therefore confirmed the prediction.

Prediction 2b was that older children would produce more translation equivalents than younger children. We observed a positive correlation between age (our proxy for LEARNABLE) and number of translation equivalents in our dataset, _r_(`r cor_age_TE$parameter[1]`) = `r round(cor_age_TE$estimate[1],2)`, $p `r papaja::printp(cor_age_TE$p.value[1])`$, and therefore confirmed the prediction.

Prediction 2c was that both older children and those with the least balanced vocabularies (BALANCE) would produce more dominant-language singlets (DOM-SINGLET). This pattern was confirmed by the results from our dataset, with a positive correlation between dominant-language singlets (DOM-SINGLET) and age (which determined LEARNABLE), _r_(`r cor_age_DOMSING$parameter[1]`) = `r round(cor_age_DOMSING$estimate[1],2)`, $p `r papaja::printp(cor_age_DOMSING$p.value[1])`$, and a negative correlation between BALANCE and dominant-language singlets (DOM-SINGLET), _r_(`r cor_BALANCE_DOMSING$parameter[1]`) = `r round(cor_BALANCE_DOMSING$estimate[1],2)`, $p `r papaja::printp(cor_BALANCE_DOMSING$p.value[1])`$. As illustrated in Figure 2 Panel 2B, children with the least balanced vocabulary of 0.1 produced the most singlets in their dominant language (mean = `r round(mean_DOMSINGLET_BALANCEsubset$mean_DOM_SINGLET[1],0)`), whereas those with the most balanced vocabulary of 0.5 produced the fewest (mean = `r round(mean_DOMSINGLET_BALANCEsubset$mean_DOM_SINGLET[5],0)`).

Prediction 2d was that older children and those with the most balanced vocabularies (BALANCE) would produce more singlets in their non-dominant language. This pattern was also observed in our dataset, with a positive correlation between the number of non-dominant singlets (NONDOM-SINGLET) and age (which determined LEARNABLE), _r_(`r cor_age_NONDOMSING$parameter[1]`) = `r round(cor_age_NONDOMSING$estimate[1],2)`, $p = `r papaja::printp(cor_age_NONDOMSING$p.value[1])`$, and a positive correlation between BALANCE and the number of non-dominant singlets (NONDOM-SINGLET), _r_(`r cor_BALANCE_NONDOMSING$parameter[1]`) = `r round(cor_BALANCE_NONDOMSING$estimate[1],2)`, $p `r papaja::printp(cor_BALANCE_NONDOMSING$p.value[1])`$. As shown in Figure 2 Panel 2C, we observed that children produced very few singlets in their non-dominant language, although the most balanced children with BALANCE of 0.5 produced the most singlets in their non-dominant language (mean = `r round(mean_NONDOMSINGLET_BALANCEsubset$mean_NONDOM_SINGLET[5],0)`) and the least balanced children with BALANCE of 0.1 produced the fewest (mean = `r round(mean_NONDOMSINGLET_BALANCEsubset$mean_NONDOM_SINGLET[1],0)`).

## 2.2.4 Testing Prediction Set 3: Rate of translation equivalent learning
Prediction Set 3 pertained to the overall nature of translation equivalent learning, describing expected patterns of translation equivalent learning under the Neutral Account, the Avoidance Account, or the Preference Account. To directly test the correspondence of our data with these different accounts, we built a linear regression model predicting the observed number of translation equivalents from the Bilingual Vocabulary Model using the formula TE = DOM×NONDOM/LEARNABLE, and we allowed the model to estimate the BIAS parameter.

```{r percentile_info, include = FALSE, cache = TRUE}
# read in the percentile file
ws_percentile <- read.csv(here::here("data_keepers/ws_percentile.csv")) %>%
  # filter out age ranges that is smaller than 18 months
  filter(age_months_percentile >= 18) %>%
  # as the upper age limit of the CDI-WS is 30m, create a new row for age 30-33m
  mutate(age_months_percentile = case_when(age_months_percentile == 30 ~ "30 - 33", 
                                           TRUE ~ as.character(age_months_percentile))) %>%
  # select only variables relevant for the 90th percentile
  select(age_months_percentile, EngWS_90percentile, FrWS_90percentile, Average_90percentile) %>%
  # rename variables for presentation
  rename("Age (months)" = age_months_percentile,
         "Number of English words produced at 90th percentile" = EngWS_90percentile,
         "Number of French words produced at 90th percentile" = FrWS_90percentile,
         "Average (LEARNABLE)" = Average_90percentile) %>%
  # round to 1 decimal place
  mutate(across(where(is.numeric),round,1))
```

First, we will walk through the parameters in this model. The size of dominant vocabulary (DOM) and size of non-dominant vocabulary (NONDOM) were taken to be the number of words produced by individual children observed in the vocabulary data. Again, the number of learnable words (LEARNABLE) was determined by averaging English and French productive CDI vocabulary at the 90th percentile at different ages (percentile information obtained from Wordbank version 0.3.1; Frank et al., 2016). Table 7 lists the denominators derived from averaging these percentiles, or LEARNABLE, at different ages. For example, for an 18 month-old infant, the denominator was `r round(ws_percentile$"Average (LEARNABLE)"[1],1)` words which was calculated by averaging the `r round(ws_percentile$"Number of English words produced at 90th percentile"[1],1)` English words and `r round(ws_percentile$"Number of French words produced at 90th percentile"[1],1)` French words, based on what 18-month-old children would typically produce at the 90th percentile. For children who were between 31 to 33 months in our dataset, the 90th percentile of 30-month-old children was used since 90th percentile information was available only up to 30 months.

```{r table7, results = "asis", cache = TRUE}
kable(ws_percentile,
        format = "latex", booktabs = TRUE, align = "c",
        caption = "Table 7. The number of total English and French productive CDI vocabulary at the 90th percentile at different ages, and the average between the two which serves as the denominator in our computation model.") %>%
  kableExtra::kable_styling(font_size = 10) %>%
  column_spec(1, width = "75px") %>%
  column_spec(2:4, width = "125px")
```

@@

## creating predicted TEs for simulation
keepers_ws_TE_simulation <- keepers_ws_TE %>%
  mutate(Predicted_TE_percentile = total_words_dom * total_words_nondom/Average_90percentile) %>%
  select(baby_id, ID_testdate, Predicted_TE_percentile, number_of_te, Average_75percentile, Average_90percentile, total_words_dom, total_words_nondom, 
         word_vocab, age_months_binned)

## linear regression model (90th percentile)
test_prediction3_model <- keepers_ws_TE_simulation %>%
  lm(number_of_te*(Average_90percentile) ~ 0 + total_words_dom:total_words_nondom, data = .)

## linear regression model (75th percentile for footnote 4)
test_prediction3_model_75percentile <- keepers_ws_TE_simulation %>%
  lm(number_of_te*(Average_75percentile) ~ 0 + total_words_dom:total_words_nondom, data = .)

@@

Our model showed an excellent model fit of $R^2$ = `r round(summary(test_prediction3_model)$r.squared,2)`, indicating that our model explained `r round(summary(test_prediction3_model)$r.squared,2) * 100`% of the variance in bilinguals’ translation equivalent knowledge. The linear regression model estimated a BIAS coefficient of `r round(summary(test_prediction3_model)$coefficient[1],2)`, $p `r papaja::printp(summary(test_prediction3_model)$coefficient[4])`$. This value is extremely close to 1, suggesting that our data are consistent with the account whereby translation equivalents are learned equivalently to other words[^4].

[^4]: We also explored how these results would vary if we set the number of LEARNABLE words at the 75th percentile, rather than at the 90th percentile. Overall, model fit was slightly reduced compared to our main model, although fit was still good, $R^2$ = `r round(summary(test_prediction3_model_75percentile)$r.squared,2)`, $p `r papaja::printp(summary(test_prediction3_model_75percentile)$coefficient[4])`$. The estimated BIAS parameter was lower at `r round(summary(test_prediction3_model_75percentile)$coefficient[1],2)`. This change in the BIAS parameter is mathematically expected due to the reciprocal nature of LEARNABLE and BIAS, an issue we will return to in the discussion.

To illustrate the close fit between the Neutral Account and our data, we used the Bilingual Vocabulary Model formula TE = 1 × (DOM×NONDOM/LEARNABLE) to estimate each child’s expected translation equivalent knowledge (setting BIAS = 1; also setting the intercept at 0 to replicate our linear regression model), which is plotted against our observed data in Figure 5. Expected and observed translation equivalents were closely aligned with the Neutral Account of the Bilingual Vocabulary Model (i.e., BIAS = 1), suggesting that the Neutral Account provides a parsimonious explanation for bilinguals’ translation equivalent knowledge. This provides evidence for the notion that translation equivalents are neither harder nor easier to learn than singlets in bilingual vocabulary learning. Note that visual inspection suggested that there could be some possible outliers. Cook’s distance was estimated for our linear regression model listed above and identified `r words(length(number_of_outliers$ID_testdate))` data points with a cook’s distance over 0.4. After removing those `r words(length(number_of_outliers$ID_testdate))` data points, the linear regression model returned a coefficient of `r round(summary(model_remove_outliers)$coefficient[1],2)`, $p `r papaja::printp(summary(model_remove_outliers)$coefficient[4])`$, with $R^2$ = `r round(summary(model_remove_outliers)$r.squared,2)`. As the model fit was similar to the model without eliminating the `r words(length(number_of_outliers$ID_testdate))` outlier data points, we proceeded with the full data set keeping the `r words(length(number_of_outliers$ID_testdate))` potential outlier data points.

@@

Despite the good overall fit to the data, a close examination of Figure 5 suggested that the model might less closely fit the data of children with smaller vocabulary sizes. Figure 7 displays the model fit separately for children with a word vocabulary (WORD) less than 300 words and those with a word vocabulary (WORD) of 300 or greater. Based on visual inspection, the slope of translation equivalent learning appeared steeper for children with less than 300 total vocabulary, suggesting that translation equivalents are more easily learned (i.e., BIAS > 1); whereas the slope of translation equivalent learning appeared to align with the Neutral Account of the Bilingual Vocabulary Model (i.e., BIAS = 1) for children with more than 300 total vocabulary. To further explore this pattern, we ran the same linear regression twice, separately for children with less than 300 total vocabulary and for those with more than 300 total vocabulary. The model for those with larger total word vocabulary (WORD) returned a coefficient of BIAS = `r round(summary(percentile_model_more300w)$coefficient[1],2)`, $p `r papaja::printp(summary(percentile_model_more300w)$coefficient[4])`$, whereas the model for those with less than 300 total word vocabulary (WORD) returned a coefficient of BIAS = `r round(summary(percentile_model_less300w)$coefficient[1],2)`, $p `r papaja::printp(summary(percentile_model_less300w)$coefficient[4])`$. Both models fit well, although a somewhat better fit was obtained for children with larger vocabulary size ($R^2$ = `r round(summary(percentile_model_more300w)$r.squared,2)`) than children with smaller vocabulary size ($R^2$ = `r round(summary(percentile_model_less300w)$r.squared,2)`). Overall, this analysis suggests that translation equivalent learning for children with larger vocabularies corresponds best to the Neutral Account, but translation equivalent learning for children with smaller vocabularies corresponds best to the Preference Account.
